---
title: "lichens_on_old_trees"
format: html
editor: visual
warning: false
echo: false
---

# Introduction

In this section we examine the impact of the age of old trees on the alpha diversity of epiphytic lichens while controlling for confounding variables such as tree size and environmental variables. To answer this question, we used two complementary approaches. The first approach look at species richness and age from the entire sample of the first two meters of the sample trees. Consequently, approach A enabled us to untangle the role of tree age versus size, and how productivity may play into this. However, to further control for size we also inventoried a fixed area each sample tree to make them comparable across different tree sizes. To find the species density All trees had in addition two subplots, one in the south and one in the north.

To further understand the causal effect of age on the alpha diversity of epiphytic lichens we will use a framework of hypotheses with multiple steps looking like this:

### Hypothesis framework

+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Hypothesis framework                                                                                                                                                      |
+===========================================================================================================================================================================+
| **First level: Is tree age related to alpha biodiversity**                                                                                                                |
|                                                                                                                                                                           |
| Null hypothesis H0: No effect of tree age on lichen alpha diversity.                                                                                                      |
|                                                                                                                                                                           |
| H~a1~: There is a positive relationship between tree age and lichen richness                                                                                              |
|                                                                                                                                                                           |
| If H~0~ falsified because species richness increases with age then we go to the next step                                                                                 |
+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Second level: Why could tree age related to increased alpha diversity**                                                                                                 |
|                                                                                                                                                                           |
| Alternative hypothesis H~a2~: Alpha diversity increases with age because tree size increases with age, which in turn increases the surface for establishment of lichens). |
|                                                                                                                                                                           |
| Prediction 2: Species richness increases more with the size than with the age of trees.                                                                                   |
|                                                                                                                                                                           |
| The next hypotheses are tested on the subplots, thus, this is species density                                                                                             |
|                                                                                                                                                                           |
| H~a3~ : Species density on the tree stem increases with age because of a longer time span for colonization.                                                               |
|                                                                                                                                                                           |
| Prediction 3: Species density increases with tree age.                                                                                                                    |
|                                                                                                                                                                           |
| Alternative hypothesis H~a4~: Species density on the tree stem increases with age because growth stagnation in older trees facilitate establishment of lichens.           |
|                                                                                                                                                                           |
| Prediction 4: Species density is more correlated with the average growth rate of the trees than their age.                                                                |
+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

# data cleaning and wrangling

#### finding the directory

```{r}
here::here()

```

#### Packages

```{r}

library(tidyverse)
library(readxl)
library(lme4)
library(here)
library(corrplot)
library(Hmisc)
source("fun_lichens_old_trees.R")
```

#### Species data import and formatting whole tree

```{r}


#importing the lichen data
lichen.data <- read_excel("lichen_data.xlsx", range = "B5:R31316")


# replacing in NA values in the numeric and logical variables####
lichen.data |>
  dplyr :: select(6:12) |>
  mutate(
    across(everything(), ~replace_na(.x, 0))
  )   -> data.subset

lichen.data[,6:12] <- data.subset


# finding the lichen species richness for each tree#####
lichen.data |>
  group_by(tree_number,locality,datatype, muncipality) %>%
  count(individuals > 0) |>
filter(`individuals > 0` == TRUE) |>
  arrange(locality,tree_number) -> lichen.data_richness


```

#### Environmental variables import and formatting

```{r}


lichenoldtrees_treedata <- readxl::read_xlsx( "lichenoldtrees_treedata_corrected.xlsx", 
                                      range = "B7:AS407")

lichenoldtrees_treedata <-format_env_data(lichenoldtrees_treedata)

```

#### Combining the species data with the environmental variables

```{r}

  
  # remember that all the column names have to match
  joined.fielddata <- full_join(lichenoldtrees_treedata, lichen.data_richness )
  
  joined.fielddata |>
    mutate(species = n) |>
    mutate(locality =factor(locality)) |>
    filter(!is.na(species)) -> species_richness_age

# making one dataframe for each tree species
species_richness_age |>
  filter(!tree_species == "spruce")  -> pine_species_richness_age

species_richness_age |>
  filter(!tree_species == "pine") -> spruce_species_richness_age
```

### Descriptive statistics of variables

There are three types of explanatory variables here, all of which have been recorded at sample tree level

Age: Tree age and bark structure ( ordinal 1-4)

Size: DBH and thickness of the the thickest branch. Categorical for pine 0-5, 5-10, 10-20, \> 20

Productivity: Elevation, basal area around each sample tree and site index H40

#### Pine

```{r}
  pine_species_richness_age<- pine_species_richness_age |> 
select(species,age,bark_structure,dbh,branch_thickness,site_index,basal_area,elevation)|> 
mutate(branch_thickness = ordered(branch_thickness))
  
  summary(pine_species_richness_age)
```

#### Spruce

```{r}
  
spruce_species_richness_age <- spruce_species_richness_age |> 
select(species,age,bark_structure,dbh,branch_thickness.numeric,site_index,basal_area,elevation)
  
  summary(spruce_species_richness_age)
```

# Analysis

## Exploratory analyses

### Kendall's tau correlation between environmental variables + species richness

**Comments to the correlation plots:**

X signifies that the correlation has a p value \> 0.05

NB: Beware of Simpson's paradox when interpreting these rank based correlation plots.

#### Pine

```{r}
pine_var <- pine_species_richness_age |>
  mutate(branch_thickness = as.ordered(branch_thickness)) |> 
  mutate(branch_thickness = as.numeric(branch_thickness)) |> 
  filter(!is.na(age))

#pine_var <- apply(pine_var,2,as.numeric) 


# correlation matrix
 pine_cor <- cor(pine_var, method = "kendall")

 #Significance values 
 pval <- cor.mtest(pine_cor) 
 corrplot( pine_cor, p.mat =  pval$p, diag = T, type = "upper", sig.level = 0.05)
```

#### Spruce

```{r}
spruce_var <- spruce_species_richness_age |> 
  filter(!is.na(age))

spruce_var <- apply(spruce_var,2,as.numeric) 


# correlation matrix
 spruce_cor <-cor( spruce_var, method = "kendall")

 #Significance values 
 pval <- cor.mtest(spruce_cor) 
 corrplot( spruce_cor, p.mat =  pval$p, diag = T, type = "upper", sig.level = 0.05)
  
```

Site index and elevation are strongly correlated with each other. Therefore, it might be wise to choose one of them later on in the alpha diversity analysis.

Branch thickness is also clearly correlated with the different productivity variables. This makes sense as low productive forests are more open which equals more light.

The lichen epiphyte richness is correlated with all the productivity variables

### exploratory age plot

The scale of the plots are the same for both tree species.

#### Pine exploratory age plot

```{r}
  
pine_species_richness_age.plot <- exploratory_plot(pine_species_richness_age) 

# ggsave("pine_species_richness_age.plot.jpeg",pine_species_richness_age.plot, dpi = 250)
# 15.9 x 11.3 brukt tidligere


print(pine_species_richness_age.plot)
```

#### Spruce exploratory age plot

```{r}

spruce_species_richness_age.plot <- exploratory_plot(spruce_species_richness_age)

print(spruce_species_richness_age.plot)
```

### Both tree species in the same plot

```{r}
scatterplot_output <- ggplot(species_richness_age,aes(y= species, x = age, color = tree_species)) +
  scale_color_manual(values = c("goldenrod1", "darkblue")) +
    geom_point(size = 2, alpha = 0.7) +
    geom_smooth(method = "lm") +
    guides(color = guide_legend(title = "tree species")) +
    theme(plot.margin=grid::unit(c(0,0,0,0), "mm")
          , axis.line = element_line(colour = "black", size = 1)
          , axis.ticks.length=unit(.25, "cm")
          ,  panel.background = element_blank()
          , axis.title.y = element_text(size = 22, face = "bold")
          , axis.text.y = element_text(size = 20)
          , axis.text.x = element_text(size = 20)
          , axis.title.x = element_text(size = 22, face = "bold" )
          , plot.title = element_text(size = 14)
          , legend.text = element_text(size = 12)
          , legend.title = element_text(size = 14, face = "bold")
          , legend.position = c(0.15,0.85)) +
    xlim(0,425) +
    ylim(0,30) 
print(scatterplot_output)
```

## Models species richness on the entire sample trees

#### First level

Testing the following hypothesis:

Null hypothesis, H0s: No effect of tree age on lichen alpha diversity.

H1: There is a positive relationship between tree age and lichen richness

```{r}

pinemod<- glmer(species~ 
                  scale(age, center = T, scale = T) 
                + (1|locality/plot), family = ("poisson"),
                data = pine_species_richness_age)


sprucemod <- glmer(species~ scale(age, center = T, scale = T)
                    + (1|locality/plot) , family = ("poisson") , data = spruce_species_richness_age)


sjPlot:: tab_model(sprucemod,pinemod,  dv.labels = c("Spruce species richness", "Pine species richness"))

```

The lichen epiphyte richness is positively related to tree age for both species, and thus the first null hypothesis is rejected. This association has the strongest evidence for spruce.There appears to be a quite small effect size for both species.

We go on to the next level in the hypothesis testing

#### **Second level: Why could tree age related to increased alpha diversity?**

Alternative hypothesis H~a2~: Alpha diversity increases with age because tree size increases with age, which in turn increases the surface for establishment of lichens).

Prediction 2: Species richness increases more with the size than with the age of trees.

```{r}
#####

pinemod<- glmer(species~ 
                  scale(age, center = T, scale = T) 
                  + scale(dbh, center = T, scale = T)
                  + branch_thickness
                 + + (1|locality/plot) , family = ("poisson") , data = pine_species_richness_age)

sprucemod <- glmer(species~ scale(age, center = T, scale = T) 
                      + scale(dbh, center = T, scale = T)
                      + scale(branch_thickness.numeric, center = T, scale = T)
                    + + (1|locality/plot) , family = ("poisson") , data = spruce_species_richness_age)

sjPlot:: tab_model(sprucemod,pinemod,  dv.labels = c("Spruce species richness", "Pine species richness"))
```

The null hypothesis cannot rejected for spruce, but none of the variables are significant for pine. The species richness seems to more or less equal due tree age and size to tree age. It is also clear from the analysis that alpha diversity is related to large branched trees for spruce trees and not necessarily just large diameter trees. This is likely due to all the pendulous lichens that grow on the branches.

This will be further tested later from the sub-plots on the stem.

#### Testing productivity

```{r}
#####

pinemod<- glmer(species~ 
                  scale(age, center = T, scale = T) 
                  + scale(site_index, center = T, scale = T)
                  + scale(elevation, center = T, scale = T)
                  + scale(basal_area, center = T, scale = T)
                + (1 |locality/plot), family = ("poisson"),
                data = pine_species_richness_age)


sprucemod <- glmer(species~ scale(age, center = T, scale = T) 
                     + scale(site_index, center = T, scale = T)
                      + scale(elevation, center = T, scale = T)
                      + scale(basal_area, center = T, scale = T)
                    + (1|locality/plot) , family = ("poisson") , data = spruce_species_richness_age)


sjPlot:: tab_model(sprucemod,pinemod,  dv.labels = c("Spruce species richness", "Pine species richness"))
```

#### Testing everything at once

```{r}

pinemod<- glmer(species~ 
                  scale(age, center = T, scale = T) 
                  + scale(site_index, center = T, scale = T)
                  + scale(dbh, center = T, scale = T)
                  + branch_thickness
                 + scale(elevation, center = T, scale = T)
                  + scale(basal_area, center = T, scale = T)
                + (1 |locality/plot) , family = ("poisson") , data = pine_species_richness_age)


sprucemod <- glmer(species~ scale(age, center = T, scale = T) 
                      + scale(site_index, center = T, scale = T)
                      + scale(dbh, center = T, scale = T)
                      + scale(branch_thickness.numeric, center = T, scale = T)
                      + scale(elevation, center = T, scale = T)
                      + scale(basal_area, center = T, scale = T)
                    + (1 |locality/plot) , family = ("poisson") , data = spruce_species_richness_age)


sjPlot:: tab_model(sprucemod,pinemod,  dv.labels = c("Spruce species richness", "Pine species richness"))


fullmod<- glmer(species~ 
                  scale(age, center = T, scale = T) + tree_species + scale(dbh, center = T, scale = T) + scale(elevation, center = T, scale = T)
                + (1|locality/plot) , family = ("poisson"),
                data = species_richness_age)
sjPlot:: tab_model(fullmod,  dv.labels = c("Species richness for a model with both trees"))
```

### Comments to the models

All of the models converge without any problems. The only error messages are from the spruce models has singular random effects structure, but otherwise no error messages.

## size independent inventory

Importing the species data from the plots

```{r}
#importing the lichen data####
lichen.data <- read_excel("lichen_data.xlsx", range = "B5:R31316")

# replacing in NA values in the numeric and logical variables####
lichen.data |>
  dplyr :: select(6:16) |>
  mutate(
    across(everything(), ~replace_na(.x, 0))
  )   -> data.subset

lichen.data[,6:16] <- data.subset



# finding the lichen species richness for each tree#####
lichen.data |>
  mutate(sub.plot = as.numeric(southside) + as.numeric(northside)) |>
  mutate(sub.plot = as.numeric(sub.plot > 0)) |>
   group_by(tree_number,locality,datatype, muncipality) %>%
  summarise( sum(sub.plot)) |>
  arrange(locality,tree_number)  -> lichen.data_richness.sub.plot
```

### Subplot figures

NB: The y axis is reduced to 14 to make it possible to see the pattern

```{r}

#importing the tree data

 
# remember that all the column names have to match
joined.fielddata <- full_join(lichenoldtrees_treedata, lichen.data_richness.sub.plot)


species_richness_age <- joined.fielddata |> 
  mutate(species = `sum(sub.plot)`)


species_richness_age |>
  filter(tree_species == "pine") -> pine_species_richness_age

species_richness_age |>
  filter(tree_species == "spruce")  -> spruce_species_richness_age



pine_species_richness_age.subplot <- exploratory_plot(pine_species_richness_age )
 
pine_species_richness_age.subplot +
  ylim(0,14)
#ggsave("pine_species_richness_age.subplot,.jpeg",pine_species_richness_age.subplot, dpi = 250)
# 15.9 x 11.3 brukt tidligere


print(pine_species_richness_age.subplot)

spruce_species_richness_age.subplot <- exploratory_plot(spruce_species_richness_age)

spruce_species_richness_age.subplot + 
  ylim(0,14)
#ggsave("spruce_species_richness_age.subplot.jpeg",spruce_species_richness_age.subplot, dpi = 250)

print(spruce_species_richness_age.subplot)
```

### Testing H~a3:~Species density on the tree stem increases with age because of a longer time span for colonization. 

If H0 falsified because alpha diversity increases with age then we move on to the species density where the area surveyed is the same.

H~a3~ : Species density on the tree stem increases with age because of a longer time span for colonization.

Prediction 3: Species density increases with tree age.

```{r}
#pinemodel
pine_species_richness_age |>
  filter(!is.na(locality )) |> 
mutate(growthrate = dbh/age)-> pine_species_richness_age


pinemod_size_independent  <- glmer(species~  
                      scale(age, center = T, scale = T)
                    + (1|locality/plot) , family = poisson,
                data = pine_species_richness_age)


#spruce model
spruce_species_richness_age|>
mutate(growthrate = dbh/age) |> 
  filter(!is.na(locality )) -> spruce_species_richness_age

sprucemod_size_independent <- glmer(species~ scale(age, center = T, scale = T)
                      + (1|locality) #+  (1|plot)
                      , family = poisson,
                data = spruce_species_richness_age)


sjPlot:: tab_model(sprucemod_size_independent, pinemod_size_independent,dv.labels = c("Spruce species density", "Pine species density") )


#both tree species model
species_richness_age|>
mutate(growthrate = dbh/age) |> 
  filter(!is.na(locality )) -> species_richness_age

full_mod_size_independent <- glmer(species~ scale(age, center = T, scale = T) + tree_species        + (1|locality/plot) 
                      , family = poisson,
                data = species_richness_age)

sjPlot:: tab_model(full_mod_size_independent, dv.labels = c("Both tree species  density") )
```

The null hypothesis is rejected for pine, but not spruce. The species density increases with tree age for pine and we accept H~a3~ .

### Testing H~a4~: Species density on the tree stem increases with age because growth stagnation in older trees 

H~a4~: Species density on the tree stem increases with age because growth stagnation in older trees facilitate establishment of lichens.

Prediction 4: Species density is more correlated with the average growth rate of the trees than their age.

```{r}

#pinemodel
pine_species_richness_age |>
  filter(!is.na(locality )) -> pine_species_richness_age


pinemod_size_independent_growth_rate  <- glmer(species~  
                      scale(growthrate, center = T, scale = T)
                    + (1|locality/plot), family = poisson , data = pine_species_richness_age)


#spruce model

sprucemod_size_independent_growth_rate <- glmer(species~ scale(growthrate, center = T, scale = T)
                      + (1|locality/plot),family = poisson, data = spruce_species_richness_age)

sjPlot:: tab_model(sprucemod_size_independent_growth_rate ,pinemod_size_independent_growth_rate ,dv.labels = c("Spruce species density", "Pine species density") )


full_mod_size_independent <- glmer(species~ scale(growthrate, center = T, scale = T) + tree_species        + (1|locality/plot) 
                      , family = poisson,
                data = species_richness_age)

sjPlot:: tab_model(full_mod_size_independent, dv.labels = c("Both tree species  density") )
```

This might be because they have more open crowns that are not shading the stem. It may be that a longer time span for colonization is more important than bark stability for pine. This makes sense as pine trees develop large bark plates that very slippery and can shed.

### Extra information

#### Kendall's tau correlation between environmental variables + species density

##### Pine

```{r}
pine_var <- pine_species_richness_age |>
  select(species, age,bark_structure,dbh, branch_thickness, site_index,elevation, basal_area) |>  
  mutate(branch_thickness = as.ordered(branch_thickness)) |> 
  mutate(branch_thickness = as.numeric(branch_thickness)) |> 
  filter(!is.na(age))

#pine_var <- apply(pine_var,2,as.numeric) 


# correlation matrix
 pine_cor <- cor(pine_var, method = "kendall")

 #Significance values 
 pval <- cor.mtest(pine_cor) 
 corrplot( pine_cor, p.mat =  pval$p, diag = T, type = "upper", sig.level = 0.05)
  
```

##### Spruce

```{r}
spruce_var <- spruce_species_richness_age |>
  select(species, age,bark_structure,dbh, branch_thickness, site_index,elevation, basal_area) |>  
  filter(!is.na(age)) |> 
  filter(!is.na(species))

spruce_var <- apply(spruce_var,2,as.numeric) 


# correlation matrix
 spruce_cor <-cor( spruce_var, method = "kendall")

 #Significance values 
 pval <- cor.mtest(spruce_cor) 
 corrplot( spruce_cor, p.mat =  pval$p, diag = T, type = "upper", sig.level = 0.05)
  
```
